\textbf{Selective DNA and Protein Isolation from Marine Macrophyte Surfaces}

### 2021-02-09 submission/manuscript.tex
Text formatting.
Fromat text.

Surfaces of marine macrophytes are colonized by a diverse microbial
community whose structure and function are poorly understood (Egan
\emph{et al.}, 2013). As less than 1 \% of all prokaryotic species are
culturable, molecular methods such as 16S rRNA sequencing, metagenomics
and metaproteomics are indispensable to study these organisms (Amann
\emph{et al.}, 1995; Su \emph{et al.}, 2012). Applying these techniques
requires an initial isolation step with the purpose of obtaining high
quality DNA and proteins.

Biological material (i.e., proteins and DNA) from pelagic microbial
communities is usually isolated by collecting cells onto filters and
subsequently isolating the target organisms or communities (Gilbert
\emph{et al.}, 2009). If a specific microbial size fraction is aimed
sequential filtration is applied (Massana \emph{et al.}, 1997; Andersson
\emph{et al.}, 2010). In contrast, obtaining microorganisms associated
to surfaces require either a cell detachment procedure prior to
isolation or the host material is co-extracted with the target material.
Methods for separating microbial cells from the host include shaking of
host tissue (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping of macrophyte surfaces (Uku \emph{et al.}, 2007) or applying
ultrasonication (Weidner \emph{et al.}, 1996; Cai \emph{et al.}, 2014).
It was shown that shaking alone is not sufficient to remove microbial
cells from surfaces, at least not from plant root surfaces
(Richter-Heitmann \emph{et al.}, 2016). Manual separation methods, such
as scraping and brushing are time consuming and subjective, as the
detachment efficiency depends on host tissue and the person performing
the procedure (Cai \emph{et al.}, 2014). Ultrasonication was proposed as
an alternative method as it is providing better results in terms of
detachment efficiency (Cai \emph{et al.}, 2014; Richter-Heitmann
\emph{et al.}, 2016). The downside of this procedure is that complete
cell removal is still not obtained and tissue disruption was observed
especially after the application of probe ultrasonication
(Richter-Heitmann \emph{et al.}, 2016). An alternative to these cell
detachment procedures is the isolation of target epiphytic compounds
together with host material (Staufenberger \emph{et al.}, 2008; Jiang
\emph{et al.}, 2015). This procedure can lead to problems in the
following processing steps such as mitochondrial and chloroplast 16S
rRNA sequence contaminations from the host (Longford \emph{et al.},
2007; Staufenberger \emph{et al.}, 2008). In addition, when performing
metagenomics and metaproteomics host material can cause biased results
towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the target
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed followed by the
isolation procedure, omitting host material contaminations. To our
knowledge, the only procedure describing a direct and selective
epiphytic DNA isolation from the surfaces of marine macrophytes was
described by Burke \emph{et al.} (2009). In contrast to previously
described methods, this protocol enables an almost complete removal of
the surface community. It was used for 16S rRNA gene clone library
construction (Burke and Thomas \emph{et al.}, 2011) and metagenome
sequencing (Burke and Peter Steinberg \emph{et al.}, 2011). This method,
although providing a selective isolation procedure, uses a rapid
multi-enzyme cleaner (3M) that is not available worldwide and the
chemical constituents are unknown (Burke \emph{et al.}, 2009). Also to
our knowledge, no selective isolation protocol to perform
(meta)proteomics of epiphytic communities associated with marine
macrophytes has been developed yet.

In the present study, we adapted a protocol widely used for DNA
isolation from filters (Massana \emph{et al.}, 1997) and a protocol used
for protein isolation from soils (Chourey \emph{et al.}, 2010; Hultman
\emph{et al.}, 2015). These two adapted methods allowed for a selective
extraction of DNA and proteins from epiphytic communities inhabiting the
surfaces of two marine macrophytes, the seagrass \emph{Cymodocea nodosa}
and the macroalga \emph{Caulerpa cylindracea}. In addition, we tested
the removal efficiency of the protocol and the suitability of obtained
DNA and proteins for 16S rRNA sequencing, metagenomics and
metaproteomics.

@@

To test whether the developed DNA and protein isolation protocols
efficiently detach microbes from the macrophyte surface we selected
\emph{C. nodosa} and \emph{C. cylindracea} as representatives of
seagrass and macroalgal species. These species differ morphologically.
While \emph{C. nodosa} leaves are flat, \emph{C. cylindracea} thalli are
characterized by an uneven surface (Kuo and den Hartog, 2001; Verlaque
\emph{et al.}, 2003). The developed protocol led to an almost complete
removal of epiphytic cells from the surfaces of both species comparable
to the result of Burke \emph{et al.} (2009), indicating that structural
differences do not impact the removal efficiency. In addition, isolation
protocols were tested in two contrasting seasons, as it is known that
macrophytes are harbouring more algal epiphytes during autumn and winter
(Reyes and Sansón, 2001). No differences in the removal efficiency was
observed between seasons suggesting that these protocols can be used on
macrophyte samples retrieved throughout the year. Also, no removal
differences were observed on samples derived from the same host but from
different locations.

Successful amplification and sequencing of the V4 region of the 16S rRNA
gene proved that the isolated DNA can be used to estimate the microbial
epiphytic diversity. Taxonomic groups detected in this step can also be
often found in epiphytic communities associated with other macrophytes
(Burke and Thomas \emph{et al.}, 2011; Morrissey \emph{et al.}, 2019). A
problem often encountered in studies focusing on epiphytic communities
is the presence of large proportions of chloroplast 16S rRNA sequences
in the pool of amplified molecules, especially if the epiphytic DNA was
isolated without prior selection (Staufenberger \emph{et al.}, 2008).
These sequences can derive from host chloroplasts or from eukaryotic
epiphytic chloroplast DNA. Although the proportion of obtained
chloroplast 16S rRNA sequences in our samples was substantial, they
derived almost exclusively from eukaryotic epiphytes. High proportion of
chloroplast 16S rRNA sequences in studies applying selective procedures
that include direct cellular lysis on host surfaces were observed before
(Michelou \emph{et al.}, 2013). It is possible that chloroplast-specific
sequences even in these studies originated from eukaryotic epiphytic
cells and not from host chloroplasts. Indeed, it is common during 16S
rRNA profiling of pelagic microbial communities to observe high
proportions of chloroplast sequences (Gilbert \emph{et al.}, 2009;
Korlević \emph{et al.}, 2016). In addition, a very low proportion of
chloroplast 16S rRNA sequences in samples originating from \emph{C.
cylindracea} in comparison to \emph{C. nodosa} could be explained by the
presence of three introns in the gene for 16S rRNA in some members of
the genus \emph{Caulerpa} that could hamper the amplification process
(Lam and Lopez-Bautista, 2016).

High quality DNA is also needed for metagenomics. The obtained number of
metagenomic sequences and assembly statistics were comparable to
metagenomes and metatranscriptomes derived from similar surface
associated communities (Crump \emph{et al.}, 2018; Cúcio \emph{et al.},
2018). In addition, functional annotation of predicted coding sequences
to COG functional categories showed that the obtained metagenomes can be
used to determine the metabolic capacity of surface associated
communities (Leary \emph{et al.}, 2014; Cúcio \emph{et al.}, 2018). The
proportion of coding sequences, including their RPKM, originating from
\emph{C. nodosa} metagenomes and classified as Streptophyta was low
indicating that the isolation procedure was specific for epiphytic
cells. DNA samples isolated from the surface of \emph{C. cylindracea}
exhibited a low proportion of Chlorophyta coding sequences, however,
their RPKM was higher than in the samples originating from \emph{C.
nodosa}. One of the reasons for this elevated RPKM of Chlorophyta
sequences in \emph{C. cylindracea} could be the differences in the
tissue structure between these two host species. While \emph{C. nodosa}
leaves are composed of individual cells, the thallus of \emph{C.
cylindracea} is, like in other siphonous algal species, composed of a
single large multinucleate cell (Coneva and Chitwood, 2015). The absence
of individual cells in \emph{C. cylindracea} could cause a leakage of
genetic material into the extraction buffer causing an elevated presence
of host sequences in the samples for metagenome analyses.

To obtain insight into the metabolic status of uncultivated prokaryotes,
a metaproteomic approach is required (Saito \emph{et al.}, 2019). The
applied protocol for epiphytic protein isolation followed by a
metaproteomic analysis identified between 14,219 and 16,449 proteins,
which is higher than previously reported for e.g.~soils (Chourey
\emph{et al.}, 2010; Hultman \emph{et al.}, 2015), seawater (Williams
\emph{et al.}, 2012) and biofilms (Leary \emph{et al.}, 2014). The
functional annotation of identified proteins into COG functional
categories showed that the protein isolation protocol can be used to
assess the metabolic status of the epiphytic community (Leary \emph{et
al.}, 2014). Similar to the results of the metagenomic analysis, the
number and abundance of identified proteins affiliated to Streptophyta
in \emph{C. nodosa} samples were low, indicating that the procedure is
selective for epiphytic cell proteins. In addition, a higher number and
abundance of identified proteins associated with Chlorophyta were
observed in \emph{C. cylindracea} samples. The cause of this elevated
presence of Chlorophyta-associated proteins can be, similar to the DNA
isolation protocol, explained by the absence of individual cells in this
siphonous alga (Coneva and Chitwood, 2015).

@@

\hypertarget{experimental-procedures}{%
\subsection{Experimental procedures}\label{experimental-procedures}}

Leaves of \emph{C. nodosa} were sampled in a \emph{C. nodosa} meadow in
the Bay of Saline, northern Adriatic Sea (\ang{45;7;5} N, \ang{13;37;20}
E) and in a \emph{C. nodosa} meadow invaded by \emph{C. cylindracea} in
the Bay of Funtana, northern Adriatic Sea (\ang{45;10;39} N,
\ang{13;35;42} E). Thalli of \emph{C. cylindracea} were sampled in the
same \emph{C. nodosa} invaded meadow in the Bay of Funtana and at a
locality of only \emph{C. cylindracea} located in the proximity of the
invaded meadow. Leaves and thalli for 16S rRNA analysis, metagenomics
and metaproteomics were collected in two contrasting seasons, on 4
December 2017 (16S rRNA analysis and metaproteomics), 14 December 2017
(metagenomics) and 18 June 2018 (16S rRNA analysis, metagenomics and
metaproteomics). During spring 2018, the \emph{C. nodosa} meadow in the
Bay of Saline decayed to an extent that no leaves could be retrieved
(Najdek \emph{et al.}, 2020). In addition, as not enough DNA for both
metagenomic and 16S RNA analysis were obtained during the sampling on 4
December 2017, an additional sampling on 14 December 2017 was carried
out in the Bay of Funtana. Leaves and thalli were collected by diving
and transported to the laboratory in containers placed on ice and filled
with seawater from this site. Upon arrival to the laboratory, \emph{C.
nodosa} leaves were cut into sections of 1 -- 2 \si{\cm}, while \emph{C.
cylindracea} thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves
and thalli were washed three times with sterile artificial seawater
(ASW) to remove loosely attached microbial cells.

The DNA was isolated according to the protocol for isolation from
filters described in Massana \emph{et al.} (1997). This protocol was
modified and adapted for microbial DNA isolation from macrophyte
surfaces as described below. Five (5) \si{\ml} of lysis buffer (40
\si{\milli\Molar} EDTA, 50 \si{\milli\Molar} Tris-HCl, 0.75 \si{\Molar}
sucrose; pH 8.3) was added to 1 \si{\g} wet weight of leaves or 2
\si{\g} wet-weight of thalli. For every sample, duplicate isolations
were performed. Lysozyme was added (final concentration 1
\si{\mg\per\ml}) and the mixture was incubated at 37 \si{\degreeCelsius}
for 30 \si{\minute}. Subsequently, proteinase K (final concentration 0.5
\si{\mg\per\ml}) and SDS (final concentration 1 \si{\percent}) were
added and the samples were incubated at 55 \si{\degreeCelsius} for 2
\si{\hour}. Following the incubation, tubes were vortexed for 10
\si{\minute} and the mixture containing lysed epiphytic cells was
separated from host leaves or thalli by transferring the solution into a
clean tube. The lysate was extracted twice with a mixture of
phenol:chloroform:isoamyl alcohol (25:24:1; pH 8) and once with
chloroform:isoamyl alcohol (24:1). After each addition of an organic
solvent mixture, tubes were slightly vortexed and centrifuged at 4,500 ×
g for 10 \si{\minute}. Following each centrifugation the aqueous phases
were retrieved. After the final extraction 1/10 of chilled 3 \si{\Molar}
sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1
volume of chilled isopropanol, incubating the mixtures overnight at
\num{-20} \si{\degreeCelsius} and centrifuging at 16,000 × g and 4
\si{\degreeCelsius} for 20 \si{\minute}. The pellet was washed twice
with 1 \si{\ml} of chilled 70 \si{\percent} ethanol and centrifuged
after each washing step at 20,000 × g and 4 \si{\degreeCelsius} for 10
\si{\minute}. After the first washing step duplicate pellets from the
same sample were pooled and transferred to a clean 1.5 \si{\ml} tube.
The dried pellet was re-suspended in 100 \si{\ul} of deionized water.

>>>>

Surfaces of marine macrophytes are colonized by a diverse microbial
community whose structure and function are poorly understood (Egan et
al., 2013). As less than 1 \% of all prokaryotic species are culturable,
molecular methods such as 16S rRNA sequencing, metagenomics and
metaproteomics are indispensable to study these organisms (Amann et al.,
1995; Su et al., 2012). Applying these techniques requires an initial
isolation step with the purpose of obtaining high quality DNA and
proteins.

Biological material (i.e., proteins and DNA) from pelagic microbial
communities is usually isolated by collecting cells onto filters and
subsequently isolating the target organisms or communities (Gilbert et
al., 2009). If a specific microbial size fraction is aimed sequential
filtration is applied (Massana et al., 1997; Andersson et al., 2010). In
contrast, obtaining microorganisms associated to surfaces require either
a cell detachment procedure prior to isolation or the host material is
co-extracted with the target material. Methods for separating microbial
cells from the host include shaking of host tissue (Gross et al., 2003;
Nõges et al., 2010), scraping of macrophyte surfaces (Uku et al., 2007)
or applying ultrasonication (Weidner et al., 1996; Cai et al., 2014). It
was shown that shaking alone is not sufficient to remove microbial cells
from surfaces, at least not from plant root surfaces (Richter-Heitmann
et al., 2016). Manual separation methods, such as scraping and brushing
are time consuming and subjective, as the detachment efficiency depends
on host tissue and the person performing the procedure (Cai et al.,
2014). Ultrasonication was proposed as an alternative method as it is
providing better results in terms of detachment efficiency (Cai et al.,
2014; Richter-Heitmann et al., 2016). The downside of this procedure is
that complete cell removal is still not obtained and tissue disruption
was observed especially after the application of probe ultrasonication
(Richter-Heitmann et al., 2016). An alternative to these cell detachment
procedures is the isolation of target epiphytic compounds together with
host material (Staufenberger et al., 2008; Jiang et al., 2015). This
procedure can lead to problems in the following processing steps such as
mitochondrial and chloroplast 16S rRNA sequence contaminations from the
host (Longford et al., 2007; Staufenberger et al., 2008). In addition,
when performing metagenomics and metaproteomics host material can cause
biased results towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the target
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed followed by the
isolation procedure, omitting host material contaminations. To our
knowledge, the only procedure describing a direct and selective
epiphytic DNA isolation from the surfaces of marine macrophytes was
described by Burke et al. (2009). In contrast to previously described
methods, this protocol enables an almost complete removal of the surface
community. It was used for 16S rRNA gene clone library construction
(Burke et al., 2011b) and metagenome sequencing (Burke et al., 2011a).
This method, although providing a selective isolation procedure, uses a
rapid multi-enzyme cleaner (3M) that is not available worldwide and the
chemical constituents are unknown (Burke et al., 2009). Also to our
knowledge, no selective isolation protocol to perform (meta)proteomics
of epiphytic communities associated with marine macrophytes has been
developed yet.

In the present study, we adapted a protocol widely used for DNA
isolation from filters (Massana et al., 1997) and a protocol used for
protein isolation from soils (Chourey et al., 2010; Hultman et al.,
2015). These two adapted methods allowed for a selective extraction of
DNA and proteins from epiphytic communities inhabiting the surfaces of
two marine macrophytes, the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. In addition, we tested the
removal efficiency of the protocol and the suitability of obtained DNA
and proteins for 16S rRNA sequencing, metagenomics and metaproteomics.

\hypertarget{materials-and-methods}{%
\subsection{Materials and methods}\label{materials-and-methods}}

Leaves of \emph{C. nodosa} were sampled in a \emph{C. nodosa} meadow in
the Bay of Saline, northern Adriatic Sea (\ang{45;7;5} N, \ang{13;37;20}
E) and in a \emph{C. nodosa} meadow invaded by \emph{C. cylindracea} in
the Bay of Funtana, northern Adriatic Sea (\ang{45;10;39} N,
\ang{13;35;42} E). Thalli of \emph{C. cylindracea} were sampled in the
same \emph{C. nodosa} invaded meadow in the Bay of Funtana and at a
locality of only \emph{C. cylindracea} located in the proximity of the
invaded meadow. Leaves and thalli for 16S rRNA analysis, metagenomics
and metaproteomics were collected in two contrasting seasons, on 4
December 2017 (16S rRNA analysis and metaproteomics), 14 December 2017
(metagenomics) and 18 June 2018 (16S rRNA analysis, metagenomics and
metaproteomics). During spring 2018, the \emph{C. nodosa} meadow in the
Bay of Saline decayed to an extent that no leaves could be retrieved
(Najdek et al., 2020). In addition, as not enough DNA for both
metagenomic and 16S RNA analysis were obtained during the sampling on 4
December 2017, an additional sampling on 14 December 2017 was carried
out in the Bay of Funtana. Leaves and thalli were collected by diving
and transported to the laboratory in containers placed on ice and filled
with seawater from this site. Upon arrival to the laboratory, \emph{C.
nodosa} leaves were cut into sections of 1 -- 2 \si{\cm}, while \emph{C.
cylindracea} thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves
and thalli were washed three times with sterile artificial seawater
(ASW) to remove loosely attached microbial cells.

The DNA was isolated according to the protocol for isolation from
filters described in Massana et al. (1997). This protocol was modified
and adapted for microbial DNA isolation from macrophyte surfaces as
described below. Five (5) \si{\ml} of lysis buffer (40 \si{\milli\Molar}
EDTA, 50 \si{\milli\Molar} Tris-HCl, 0.75 \si{\Molar} sucrose; pH 8.3)
was added to 1 \si{\g} wet weight of leaves or 2 \si{\g} wet-weight of
thalli. For every sample, duplicate isolations were performed. Lysozyme
was added (final concentration 1 \si{\mg\per\ml}) and the mixture was
incubated at 37 \si{\degreeCelsius} for 30 \si{\minute}. Subsequently,
proteinase K (final concentration 0.5 \si{\mg\per\ml}) and SDS (final
concentration 1 \si{\percent}) were added and the samples were incubated
at 55 \si{\degreeCelsius} for 2 \si{\hour}. Following the incubation,
tubes were vortexed for 10 \si{\minute} and the mixture containing lysed
epiphytic cells was separated from host leaves or thalli by transferring
the solution into a clean tube. The lysate was extracted twice with a
mixture of phenol:chloroform:isoamyl alcohol (25:24:1; pH 8) and once
with chloroform:isoamyl alcohol (24:1). After each addition of an
organic solvent mixture, tubes were slightly vortexed and centrifuged at
4,500 × g for 10 \si{\minute}. Following each centrifugation the aqueous
phases were retrieved. After the final extraction 1/10 of chilled 3
\si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by
adding 1 volume of chilled isopropanol, incubating the mixtures
overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 16,000 ×
g and 4 \si{\degreeCelsius} for 20 \si{\minute}. The pellet was washed
twice with 1 \si{\ml} of chilled 70 \si{\percent} ethanol and
centrifuged after each washing step at 20,000 × g and 4
\si{\degreeCelsius} for 10 \si{\minute}. After the first washing step
duplicate pellets from the same sample were pooled and transferred to a
clean 1.5 \si{\ml} tube. The dried pellet was re-suspended in 100
\si{\ul} of deionized water.

An aliquot of isolated DNA was treated with RNase A (final concentration
200 \si{\ug\per\ml}) for 2 \si{\hour} at 37 \si{\degreeCelsius}. The DNA
concentration was determined using the Quant-iT PicoGreen dsDNA Assay
Kit (Invitrogen, USA) according to the manufacturer's instructions and
diluted to 1 \si{\ng\per\ul}. The V4 region of the 16S rRNA gene was
amplified using a two-step PCR procedure. In the first PCR the 515F
(\(5'\)-GTGYCAGCMGCCGCGGTAA-\(3'\)) and 806R
(\(5'\)-GGACTACNVGGGTWTCTAAT-\(3'\)) primers from the Earth Microbiome
Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used to amplify the target region (Caporaso et al., 2012; Apprill
et al., 2015; Parada et al., 2016). These primers contained on their
\(5'\) end a tagged sequence. Each sample was amplified in four parallel
25 \si{\ul} reactions of which each contained 1 × Q5 Reaction Buffer,
0.2 \si{\milli\Molar} of dNTPmix, 0.7 \si{\mg\per\ml} BSA (Bovine Serum
Albumin), 0.2 \si{\micro\Molar} of forward and reverse primers, 0.5 U of
Q5 High-Fidelity DNA Polymerase (New England Biolabs, USA) and 5
\si{\ng} of DNA template. Cycling conditions were: initial denaturation
at 94 \si{\degreeCelsius} for 3 \si{\minute}, 20 cycles of denaturation
at 94 \si{\degreeCelsius} for 45 \si{\s}, annealing at 50
\si{\degreeCelsius} for 60 \si{\s} and elongation at 72
\si{\degreeCelsius} for 90 \si{\s}, finalized by an elongation step at
72 \si{\degreeCelsius} for 10 \si{\minute}. The four parallel reactions
volumes were pooled and PCR products were purified using the GeneJET PCR
Purification Kit (Thermo Fisher Scientific, USA) according to the
manufacturer's instructions and following the protocol that included
isopropanol addition for better small DNA fragment yield. The column was
eluted in 30 \si{\ul} of deionized water. Purified PCR products were
sent for Illumina MiSeq sequencing (2 × 250 bp) at IMGM Laboratories,
Martinsried, Germany. Before sequencing at IMGM, the second PCR
amplification of the two-step PCR procedure was performed using primers
targeting the tagged region incorporated in the first PCR. In addition,
these primers contained adapter and sample-specific index sequences. The
second PCR was carried out for 8 cycles. Beside samples, a positive and
negative control were sequenced. A negative control was comprised of
four parallel PCR reactions without DNA template, while for a positive
control a mock community composed of evenly mixed DNA material
originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was
used. Partial 16S rRNA sequences obtained in this study have been
deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under
accession numbers SAMEA6786270, SAMEA6648792 -- SAMEA6648794,
SAMEA6648809 -- SAMEA6648811 and SAMEA6648824.

Obtained sequences were analysed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss et al., 2009) according to the MiSeq Standard
Operating Procedure (MiSeq SOP; \url{https://mothur.org/wiki/MiSeq_SOP})
(Kozich et al., 2013) and recommendations given from the Riffomonas
project to enhance data reproducibility
(\url{http://www.riffomonas.org/}). For alignment and classification of
sequences the SILVA SSU Ref NR 99 database (release 138;
\url{http://www.arb-silva.de}) was used (Quast et al., 2013; Yilmaz et
al., 2014). Sequences classified as chloroplasts by SILVA were exported
and reclassified using mothur and the RDP
(\url{http://rdp.cme.msu.edu/}) training set (version 16) reference
files adapted for mothur (Cole et al., 2014). In comparison to SILVA,
RDP allows a more detailed classification of chloroplast sequences.
Based on the ATCC MSA-1002 mock community included in the analysis a
sequencing error rate of 0.009 \si{\percent} was determined, which is in
line with previously reported values for next-generation sequencing data
(Kozich et al., 2013; Schloss et al., 2016). In addition, the negative
control processed together with the samples yielded only 2 sequences
after sequence quality curation.

Obtained sequences were analysed on the Life Science Compute Cluster
(LiSC) (CUBE -- Computational Systems Biology, University of Vienna).
Individual sequences were assembled using MEGAHIT (version 1.1.2) (Li et
al., 2015) under default settings. Putative genes were predicted from
contigs longer than 200 bp using Prodigal (version 2.6.3) (Hyatt et al.,
2010) in metagenome mode (-p meta). Abundances of predicted genes were
expressed as Reads Per Kilobase Million (RPKM) and calculated using the
BWA algorithm (version 0.7.16a) (Li and Durbin, 2009). All predicted
genes were functionally annotated using the eggNOG-mapper (Huerta-Cepas
et al., 2017) and eggNOG database (version 5.0) (Huerta-Cepas et al.,
2019). Sequence taxonomy classification was determined using the lowest
common ancestor algorithm adapted from DIAMOND (version 0.8.36)
(Buchfink et al., 2015) and by searching against the NCBI non-redundant
database (NR). To determine the phylogeny, the top 10 \si{\percent} hits
with an e-value \textless{} 1 × 10\textsuperscript{-5} were used
(-\/-top 10). Sequence renaming, coverage information computing and
metagenomic statistics calculations were performed using software tools
from BBTools (\url{https://jgi.doe.gov/data-and-tools/bbtools}).
Metagenomic sequences obtained in this study have been deposited in the
European Nucleotide Archive (ENA) at EMBL-EBI under accession numbers
SAMEA6648795, SAMEA6648797, SAMEA6648809 and SAMEA6648811.

Proteins were isolated according to the protocol for protein isolation
from soil described in Chourey et al. (2010) and modified by Hultman et
al. (2015). This protocol was further modified and adapted for microbial
protein isolation from macrophyte surfaces as described below. Twenty
(20) \si{\ml} of protein extraction buffer (4 \si{\percent} SDS, 100
\si{\milli\Molar} Tris-HCl {[}pH 8.0{]}) was added to 5 \si{\g} wet
weight of leaves or 10 \si{\g} wet weight of thalli. The mixture was
incubated in boiling water for 5 \si{\minute}, vortexed for 10
\si{\minute} and incubated again in boiling water for 5 \si{\minute}.
After a brief vortex, the lysate was transferred to a clean tube
separating the host leaves or thalli from the mixture containing lysed
epiphytic cells. Dithiothreitol (DTT; final concentration 24
\si{\milli\Molar}) was added and proteins were precipitated with chilled
100 \si{\percent} trichloroacetic acid (TCA; final concentration 20
\si{\percent}) overnight at \num{-20} \si{\degreeCelsius}. Precipitated
proteins were centrifuged at 10,000 × g and 4 \si{\degreeCelsius} for 40
\si{\minute}. The obtained protein pellet was washed three times with
chilled acetone. During the first washing step the pellet was
transferred to a clean 1.5 \si{\ml} tube. After each washing step,
samples were centrifuged at 20,000 × g and 4 \si{\degreeCelsius} for 5
\si{\minute}. Dried pellets were stored at \num{-80} \si{\degreeCelsius}
until further analysis.

Isolated proteins were whole trypsin digested using the FASP
(filter-aided sample preparation) Protein Digestion Kit (Expedeon, UK)
according to the manufacturer's instructions with small modifications
(Wiśniewski et al., 2009). Prior to loading the solution onto the
column, protein pellets were solubilized in a urea sample buffer
included in the kit amended with DTT (final concentration 100
\si{\milli\Molar}) for 45 \si{\minute} at room temperature and
centrifuged at 20,000 × g for 2 -- 5 \si{\minute} at room temperature to
remove larger particles. The first washing step after protein solution
loading was repeated twice. In addition, the centrifugation steps were
prolonged if the column was clogged. Trypsin digestion was performed on
column filters at 37 \si{\degreeCelsius} overnight for 18 \si{\hour}.
The final filtrate containing peptides was acidified with 1
\si{\percent} (final concentration) trifluoroacetic acid (TFA), freezed
at \num{-80} \si{\degreeCelsius}, lyophilized and sent to VIME -- Vienna
Metabolomics Center (University of Vienna) for metaproteomic analysis.
Peptides were re-suspended in 1 \si{\percent} (final concentration) TFA,
desalted using the Pierce C18 Tips (Thermo Fisher Scientific, USA)
according to the manufacturer's instructions and sequenced on a Q
Exactive Hybrid Quadrupole-Orbitrap Mass Spectrometer (Thermo Fisher
Scientific, USA). Obtained MS/MS spectra were searched against a protein
database composed of combined sequenced metagenomes using SEQUEST-HT
engines and validated with Percolator in Proteome Discoverer 2.1 (Thermo
Fisher Scientific, USA). The target-decoy approach was used to reduce
the probability of false peptide identification. Results whose false
discovery rate at the peptide level was \textless{} \SI{1}{\percent}
were kept. For protein identification a minimum of two peptides and one
unique peptide were required. For protein quantification, a
chromatographic peak area-based free quantitative method was applied.

Processing and visualization of 16S rRNA, metagenomic and metaproteomic
data were done using R (version 3.6.0) (R Core Team, 2019), package
tidyverse (version 1.3.0) (Wickham et al., 2019) and multiple other
packages (Neuwirth, 2014; Xie, 2014, 2015, 2019, 2020; Wilke, 2018; Xie
et al., 2018; Allaire et al., 2019; Zhu, 2019; Bengtsson, 2020). The
detailed analysis procedure including the R Markdown file for this paper
are available as a GitHub repository
(\url{https://github.com/MicrobesRovinj/Korlevic_SelectiveRemoval_FrontMicrobiol_2021}).

@@

To test whether the developed DNA and protein isolation protocols
efficiently detach microbes from the macrophyte surface we selected
\emph{C. nodosa} and \emph{C. cylindracea} as representatives of
seagrass and macroalgal species. These species differ morphologically.
While \emph{C. nodosa} leaves are flat, \emph{C. cylindracea} thalli are
characterized by an uneven surface (Kuo and den Hartog, 2001; Verlaque
et al., 2003). The developed protocol led to an almost complete removal
of epiphytic cells from the surfaces of both species comparable to the
result of Burke et al. (2009), indicating that structural differences do
not impact the removal efficiency. In addition, isolation protocols were
tested in two contrasting seasons, as it is known that macrophytes are
harbouring more algal epiphytes during autumn and winter (Reyes and
Sansón, 2001). No differences in the removal efficiency was observed
between seasons suggesting that these protocols can be used on
macrophyte samples retrieved throughout the year. Also, no removal
differences were observed on samples derived from the same host but from
different locations.

Successful amplification and sequencing of the V4 region of the 16S rRNA
gene proved that the isolated DNA can be used to estimate the microbial
epiphytic diversity. Taxonomic groups detected in this step can also be
often found in epiphytic communities associated with other macrophytes
(Burke et al., 2011b; Morrissey et al., 2019). A problem often
encountered in studies focusing on epiphytic communities is the presence
of large proportions of chloroplast 16S rRNA sequences in the pool of
amplified molecules, especially if the epiphytic DNA was isolated
without prior selection (Staufenberger et al., 2008). These sequences
can derive from host chloroplasts or from eukaryotic epiphytic
chloroplast DNA. Although the proportion of obtained chloroplast 16S
rRNA sequences in our samples was substantial, they derived almost
exclusively from eukaryotic epiphytes. High proportion of chloroplast
16S rRNA sequences in studies applying selective procedures that include
direct cellular lysis on host surfaces were observed before (Michelou et
al., 2013). It is possible that chloroplast-specific sequences even in
these studies originated from eukaryotic epiphytic cells and not from
host chloroplasts. Indeed, it is common during 16S rRNA profiling of
pelagic microbial communities to observe high proportions of chloroplast
sequences (Gilbert et al., 2009; Korlević et al., 2016). In addition, a
very low proportion of chloroplast 16S rRNA sequences in samples
originating from \emph{C. cylindracea} in comparison to \emph{C. nodosa}
could be explained by the presence of three introns in the gene for 16S
rRNA in some members of the genus \emph{Caulerpa} that could hamper the
amplification process (Lam and Lopez-Bautista, 2016).

High quality DNA is also needed for metagenomics. The obtained number of
metagenomic sequences and assembly statistics were comparable to
metagenomes and metatranscriptomes derived from similar surface
associated communities (Crump et al., 2018; Cúcio et al., 2018). In
addition, functional annotation of predicted coding sequences to COG
functional categories showed that the obtained metagenomes can be used
to determine the metabolic capacity of surface associated communities
(Leary et al., 2014; Cúcio et al., 2018). The proportion of coding
sequences, including their RPKM, originating from \emph{C. nodosa}
metagenomes and classified as Streptophyta was low indicating that the
isolation procedure was specific for epiphytic cells. DNA samples
isolated from the surface of \emph{C. cylindracea} exhibited a low
proportion of Chlorophyta coding sequences, however, their RPKM was
higher than in the samples originating from \emph{C. nodosa}. One of the
reasons for this elevated RPKM of Chlorophyta sequences in \emph{C.
cylindracea} could be the differences in the tissue structure between
these two host species. While \emph{C. nodosa} leaves are composed of
individual cells, the thallus of \emph{C. cylindracea} is, like in other
siphonous algal species, composed of a single large multinucleate cell
(Coneva and Chitwood, 2015). The absence of individual cells in \emph{C.
cylindracea} could cause a leakage of genetic material into the
extraction buffer causing an elevated presence of host sequences in the
samples for metagenome analyses.

To obtain insight into the metabolic status of uncultivated prokaryotes,
a metaproteomic approach is required (Saito et al., 2019). The applied
protocol for epiphytic protein isolation followed by a metaproteomic
analysis identified between 14,219 and 16,449 proteins, which is higher
than previously reported for e.g.~soils (Chourey et al., 2010; Hultman
et al., 2015), seawater (Williams et al., 2012) and biofilms (Leary et
al., 2014). The functional annotation of identified proteins into COG
functional categories showed that the protein isolation protocol can be
used to assess the metabolic status of the epiphytic community (Leary et
al., 2014). Similar to the results of the metagenomic analysis, the
number and abundance of identified proteins affiliated to Streptophyta
in \emph{C. nodosa} samples were low, indicating that the procedure is
selective for epiphytic cell proteins. In addition, a higher number and
abundance of identified proteins associated with Chlorophyta were
observed in \emph{C. cylindracea} samples. The cause of this elevated
presence of Chlorophyta-associated proteins can be, similar to the DNA
isolation protocol, explained by the absence of individual cells in this
siphonous alga (Coneva and Chitwood, 2015).

@@
